---
title:  Did you know that Social Security data is used for much more than benefits computing?
author:
output: html_document
---

<style>
body {
text-align: justify}
title {
font-style: bold}
</style>

<p>&nbsp;</p>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = FALSE, include=FALSE, message=FALSE}
                            ## Basic data cleaning and organization ##

# DATA IMPORTATION:
data <- read.csv("social-security-dataset.csv")
# Data Restructuring:
data <- data[-c(1,2,3,4,77:88), ] 
colnames(data) <- c('FiscalYear','Month','PublicDatasets','RestrictedPublicDatasets','Non-PublicDatasets','TotalDatasets')
#Converting numbers to numeric
data$PublicDatasets <- as.numeric(gsub(",", "", data$PublicDatasets))
data$RestrictedPublicDatasets <- as.numeric(gsub(",", "", data$RestrictedPublicDatasets))
data$`Non-PublicDatasets` <- as.numeric(gsub(",", "", data$`Non-PublicDatasets`))
data$TotalDatasets <- as.numeric(gsub(",", "", data$TotalDatasets))
#creating 'Period' column
#install.packages("stringr")
library(stringr)
data$Month <- substr(data$Month, 1,3)
data$Period <- str_c(data$Month, ' ', data$FiscalYear)
# BIG EVENTS IN USA:
#no event in 2015 is included in the study
data$Events[data$Period == "Nov 2016"] <- c("Trump Elected US President 2016")
data$Events[data$Period == "Oct 2017"] <- c("MeToo Movement in USA")
data$Events[data$Period == "Mar 2018"] <- c("Cambridge Analytica Data Scandal Leaks")
data$Events[data$Period == "Jan 2019"] <- c("National Data Privacy Day")
data$Events[data$Period == "Jan 2020"] <- c("Covid-19 Global Pandemic")
data$Events[data$Period == "May 2020"] <- c("Black Lives Matter")
data$Events[data$Period == "Jul 2020"] <- c("TikTok US data harvest")
data$Events[data$Period == "Nov 2020"] <- c("Biden Elected US President 2020")
data$Events[data$Period == "Dec 2020"] <- c("First Covid-19 Vaccinations")
data[is.na(data)] <- "No Event"
```
  
<p>&nbsp;</p>  
## Highlights:  
>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Social Security datasets are required by different categories of entities for various purposes, now more than ever. Social Security was originally intended to only provide benefits for retired, unemployed and disadvantaged Americans. However, by now, it covers more benefits, hence the universality of its use such as health insurance inclusion and the creation of its trust fund.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Slowly, the Social Security number became the single most adopted identification number in the *United States of America* both by government agencies and the private sector. The significance of this 9-digit number translates to the importance of the data connected to it that was harvested from the *Social Security Administration* (*SSA*) of the USA. It not only involves information about the beneficiaries of the social security but also about their respective employers and level of income and much more. Besides computing benefit amounts for citizens eligible for the social insurance programs, the Social Security administrative data is relevant to private businesses, governmental as well as non-governmental organizations and also research labs where studies are conducted for policy evaluation, innovation and development. The datasets provided by the SSA fall into 3 different categories: **Public**, **Restricted Public** or **Non-Public** datasets.  

```{r echo=FALSE, message=FALSE, warning=FALSE, fig2, fig.width=10,fig.asp=0.62}
#reshaping data into %
#install.packages("dplyr")
library(dplyr)
df2 <- data[,c(1,3,4,5,6)]
df2 <- df2 %>%
  group_by(FiscalYear) %>%
  summarize(Public = 100*sum(PublicDatasets)/sum(TotalDatasets), RestrictedPublic = 100*sum(RestrictedPublicDatasets)/sum(TotalDatasets), NonPublic = 100*sum(`Non-PublicDatasets`)/sum(TotalDatasets))
#install.packages("reshape2")
library(reshape2)
df2 <- melt(df2 ,  id.vars = 'FiscalYear',variable.name = 'Datasets',value.name = 'Percentage')
df2$Percentage <- round(df2$Percentage,2)
# Creating a grouped bar graph (dodge type)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("RcolorBrewer")
library(RColorBrewer)
ggplot(data = df2, mapping = aes(x=FiscalYear, y=Percentage, fill=Datasets)) +
  geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label = Percentage), vjust = -0.2, size = 3,position = position_dodge(0.9))+
  scale_fill_brewer(palette="Set2")+
  ggtitle("\nFigure 1: Social Security Datasets' Percentages (2015-2020)\n\n")+
  theme(plot.title = element_text(size =20, face = "bold"),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(size = 0.5, linetype = "solid",
                                   colour = "black"))
```
<p>&nbsp;</p>
#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Despite the dynamic events that the *United States of America* has witnessed both on a national and global level in the previous 5 years, the Social Security Administration managed to provide a scaling number of datasets. The nature and proportion of these datasets, however, varies in time.  
#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Looking at the figure above, a noticeable pattern shows the increase in *Public* datasets' percentage from the total number datasets in the SSA from 2015 to 2020 from **41.02%** to **71.51%** and a less proportionate percentage decrease in both *Non-Public* and *Restricted Public* datasets that accounts to about **half** of the respective percentages from 2015 to 2020.  
#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;No matter the nature of the events that occurred throughout these 6 fiscal years, they make space for potential research studies on a limitless number of topics. R&D professionals tend to link Social Security data with their survey data to successfully conduct their studies. This is due to the restricted scope of variables in SSA's *public* datasets. On the other hand, *restricted public* and *non-public* datasets cover more ground in terms of features which makes them rare for the public. This is proven in the previous  chart with the largest respective percentages of **50.73%** and **8.26%** from a total of **10,791 datasets** in the year 2015.  
#### *The progressive change in these numbers poses a question regarding what happened in the US since 2016 that could have possibly stimulated this or been affected by it.*  

<p>&nbsp;</p>    
```{r echo=FALSE, message=FALSE, warning=FALSE, fig3, fig.width=11, fig.asp=0.65}
# Packages
#install.packages("plotly")
library(plotly)
#install.packages("hrbrthemes")
library(hrbrthemes)
#install.packages("viridis")
library(viridis)
#convert Period to factor
df1 <-data[data$FiscalYear != 2015,]
df1$Period <- factor(df1$Period,levels = c("Jan 2016","Feb 2016","Mar 2016","Apr 2016","May 2016","Jun 2016","Jul 2016","Aug 2016","Sep 2016","Oct 2016","Nov 2016","Dec 2016","Jan 2017","Feb 2017","Mar 2017","Apr 2017","May 2017","Jun 2017","Jul 2017","Aug 2017","Sep 2017","Oct 2017","Nov 2017","Dec 2017","Jan 2018","Feb 2018","Mar 2018","Apr 2018","May 2018","Jun 2018","Jul 2018","Aug 2018","Sep 2018","Oct 2018","Nov 2018","Dec 2018","Jan 2019","Feb 2019","Mar 2019","Apr 2019","May 2019","Jun 2019","Jul 2019","Aug 2019","Sep 2019","Oct 2019","Nov 2019","Dec 2019","Jan 2020","Feb 2020","Mar 2020","Apr 2020","May 2020","Jun 2020","Jul 2020","Aug 2020","Sep 2020","Oct 2020","Nov 2020","Dec 2020"))
#Filter data points to highlight:
df1[df1$Events == "No Event",]$Events <- "None"
df_highlight <- df1[df1$Events != "None",]
df1$Events <- factor(df1$Events,levels = c("Trump Elected US President 2016","MeToo Movement in USA",
                                           "Cambridge Analytica Data Scandal Leaks","Trump 1st Impeachement Trial",
                                           "Covid-19 Global Pandemic","Black Lives Matter","TikTok US data harvest",
                                           "Biden Elected US President 2020","First Covid-19 Vaccinations"))
# Build chart
p <- df1 %>%
  ggplot( aes(x=Period, y=PublicDatasets,group=1)) +
    geom_line(color="#69b3a2") +
    ggtitle("Figure 2: Number of Public Datasets (2016-2020)\n")+
    geom_point(data=df_highlight,aes(x=Period,y=PublicDatasets,color=Events),size=1)+
    scale_x_discrete(guide = guide_axis(n.dodge=2))+
    scale_y_continuous(limits = c(0,1500))+
    ylab("Public Datasets") +
    theme(legend.key.size = unit(2, "cm"),
          axis.text.x = element_text(angle = 90),
          plot.title = element_text(size = 15, face = "bold"),
          panel.background = element_rect(fill = "white"),
          panel.grid.major = element_line(color = "grey",size =0.5),
          axis.line = element_line(size = 0.5, linetype = "solid",
                                   colour = "black"))

# Turn it interactive with ggplotly
p <- ggplotly(p)
p %>% layout(legend = list(traceorder="reversed"))
```
<p>&nbsp;</p>
#### The chart illustrates an identical pattern in the number of public datasets from 2016 to 2020, specifically in the 4th quarter of every year. It also highlights a few alarming events for the American Citizen which could represent a probable reason for the observed pattern.  

### **The SSA minimizes data politicization during presidential elections**

### SSA restricts over 100 datasets  prior to Trump's presidential election in 2016  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;One month prior to the presidential elections of Donald Trump in 2016, the number of publicly released datasets dropped from **969** to **811**. This accounts to over **100** datasets that either have been restricted to a certain number of people or organizations or clasfified as non-public for the purpose of minimzing the risk of data exploitation and politicization for election polls or they were completely disposed of by the Social Security Administration.  

### SSA shows a decrease of 52 public datasets prior to Biden's election in 2020  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The drop in the number of public social security datasets also reoccured one month prior to the presidential elections of Joe Biden in 2020 and it accounted to a decrease of **52** datasets from September to November 2020. However, between the two previous elections, the loss gap has got slowly smaller than **100** over the years, especially after the national scandal of Cambridge Analytica Data Leaks in March 2018. Documents from Cambridge Analytica in London revealed that the firm improperly obtained and used over **87 million** Facebook user profiles in a transaction with Donald Trump's presidential compaign where the scraped private Facebook data was used to build voter profiles and assist the candidate in the US presidential elections of 2016. This national data privacy crisis proved that data exploitation is not limited to commerical purposes and that it can target the American Citizen without their consent.  

### **Social security data contributes in making the workplace safer**  

### 152 public datasets were restricted during the "Me Too" Movement in the US

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The same phenomena occured with a drop rate of **13.18%** in the beginning of the US "Me too" Movement from October to September 2017. "**Me too**" is a global movement that condemns sexual harrasement and gives a voice to its victims.  
#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;This movement was led by activists, the victims themselves as well as lawyers who offered their support with pro bono cases and needed access to more data about their clients' workplaces especially in cases of sexual harrasment in a working environment. The sudden decrease could reflect the restriction of some of these **152** public social security datasets either by choice of the SSA or by request from the respective victims' representatives.   

### **Social Security Administration reacts to data privacy**  

### SSA gives access to 1312 public datasets on Data Privacy Day in January 2019  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Subsequently, the National Data Privacy Day on January 28th, 2019 was like no other in the US. Many institutions manifested its importance and spred awareness about the cause after the Cambridge Analytica scandal, including the Social Security Administration.  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The SSA became more dedicated to protecting Personally Identifiable Information (PII) and securing the public social security datasets portal for more ethical use in the future, hence, it publicly released **1312** datasets in January 2019, a record breaking number since the start of 2016.  
<p>&nbsp;</p>
```{r echo=FALSE, message=FALSE, warning=FALSE, fig4, fig.width=9, fig.asp=0.62}
#data about 2020 events
df_2020 <- data[data$FiscalYear == "2020",c(2,3,8)]
df_2020$Month <- factor(df_2020$Month, levels = c("Jan", "Feb", "Mar","Apr","May","Jun",
                                        "Jul","Aug","Sep","Oct","Nov","Dec"))
#Bar Plot:
ggplot(data = df_2020, mapping = aes(x=Month, y=PublicDatasets, fill=Events)) +
  geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label = PublicDatasets), vjust = -0.2, size = 3,position = position_dodge(0.9))+
  scale_fill_brewer(palette="Pastel2")+
  ggtitle("\nFigure 3: Public Datasets Released in 2020\n\n")+
  theme(plot.title = element_text(size =19, face = "bold"),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(size = 0.5, linetype = "solid",
                                   colour = "black"))
```
<p>&nbsp;</p>
### **Social Security data becomes more available during Covid-19 crisis**  

### SSA's monthly public datasets reach the maximum of 1469 since the pandemic

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The year 2020 was a very unbalanced year for the world and in particular the United States of America where many events affected almost every American citizen. The Covid-19 Pandemic in January 2020 was a global health crisis that did influence how early aged employees with chronic diseases decided to retire and requested their social security from the SSA. This made an impact on the urge to get access to social security data especially by national health researchers in the USA who are trying to put an end to the **"infodemic"** during the crisis by preventing the spread of misleading or fake information and news.  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Until the 2020 presidential elections, the number of public datasets released by the SSA in 2020 has been continuously increasing from **1430** to a maximum value of **1469** in order to satisfy these research needs. Moreover, non-governmental organizations were in need of social security data in the year 2020 as they worked on advocating equal health care opportunities such as First Covid-19 Vaccinations starting from December 2020 to all US social security beneficiaries which will probably result in a new rise in the beginning of 2021.  

### **Social Security data contributes to #BLM**  

### SSA publicly releases 1457 datasets during #BLM movement  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Protests for the Black Lives Matter globel movement that was initiated in the USA shortly after the death of George Floyd in May 2020 have generated a large number of petitions and fundraising campaigns regarding #BlackLivesMatter across the internet and numerous sensibilization events about ending racism globally. The movement also approached the BLM cause on an international level which perhaps pushed the SSA to move some datasets from being restricted or non-public to being publicly available to everyone everywhere.  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;This was proven in the fiscal year 2020, given the exceptionally high number of public datasets released of **1457** in May 2020. The data was used to highlight the gap between white and black people's wages, benefits and much more with the purpose of raising awareness about the details of racism present in an american's everyday activities.  

### **TikTok App challenges data ethics values of the SSA**  

### SSA releases 1463 public datasets during TikTok US data harvest allegations  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The issue of uncertain data privacy was brought up once again with the TikTok US data harvest risks. The main concern of the American users was the possibility that history with *Facebook* could possibly repeat itself with *TikTok* after July 2020; when the number of publicly available social security datasets unusually reached **1463**. By the end of 2020, data became something extremely accessible but the concern was not about the quantity but rather the use of the data. Researchers claim that the TikTok app starts collecting all types of data the moment the app is installed on the device and utilisze it in their custom advertisments' display.  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The Social Security Administration works tirelessly to secure american citizens' private data and publicly release other data with their consent. But, what would be the point if social media giants such as TikTok can collect and share this data with third parties in a matter of seconds due to ambigiuous privacy policies and terms of service?  

<p>&nbsp;</p>  
## Review  

#### &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The Social Security Administration has been playing an active role in 2020 when it comes to endorsing public data to support Research & Development departments in different fields in the United States of America. This includes global health which is the world's priority during the Covid-19 crisis and many more fields. The data that is publicly released on a yearly basis also contributes in recovery from other types of crisis, raising awareness about data privacy threats from social media or content creation giants and about data transparency when it comes to policy making and elections in the US. As for the restrictions made on private datasets to the administration, some are working on protecting US citizens data from political exploitation. Nonetheless, some are preventing NGOs from tackling certain topics and keeping the data from speaking for the voiceless. The SSA needs to open calls for datasets where every external organization can make a data usage proposal for certain restricted or non-public datasets added by the administration. How else would social security data be as transparent as the SSA claims? How else would the world use the data for positive social change and ethical purposes?  
  
<p>&nbsp;</p>
## Data   
>
**Source**: *data.gov*  
**Publisher**: Social Security Administration  
**Last updated**: July 2nd, 2021.  
**Retrieved**: June 26th, 2021.  
**Description**: *Social Security continues to release data in support of the open data initiative. Each month it updates a progress chart to reflect new datasets added (public, restricted, non-public)*  
**URL**: *https://catalog.data.gov/dataset/enterprise-data-inventory-progress-information* 



